A formal treatment of bidirectional typing

### 2023-10-11 tex/ESOP24/S4-pre-synthesis.tex
[ update ] Add ESOP submission draft

\section{Mode Decoration and Related Properties}\label{sec:pre-synthesis}

@@

\subsection{Soundness and Completeness}
\label{sec:soundness-and-completeness}

@@

\begin{proof}
From left to right:
Induction on the bidirectional typing derivation, mapping every rule to its counterpart except when constructing a typing derivation from the $\ChkRule{Sub}$ rule, in which case the induction hypothesis $\Gamma |-_{\Sigma,\erase\Omega} t : B$ suffices due to the premise $B = A$.

From right to left:
Induction on the mode derivation.
For \SynRule{Var}, \SynRule{Anno}, and \Rule{Op}, the outermost rule used in the typing derivation must be the corresponding typing rule, so by the induction hypotheses we have bidirectional typing derivations for all the sub-terms, to which we can then apply the corresponding bidirectional typing rule.
The \ChkRule{Sub} case is similar but slightly simpler: the induction hypothesis directly gives us a derivation of $\Gamma |-_{\Sigma,\Omega} t \syn A$, to which we apply $\ChkRule{Sub}$.
\end{proof}

\subsection{Generalised Mode Decoration}
\label{sec:mode-decoration}

@@

\begin{figure}
  \centering
  \small
  \begin{tabular}{ r r l }
    & & is in mode~$d$, \\
    $\smash{\boxed{V |-_{\Sigma, \Omega} \isTerm{t}^{\dir{d}\,g\,s}}}$
    & The raw term~$\isTerm{t}$\hspace{-.6em}
    & misses some type annotation iff $g = \BFalse$, and \\
    & & is in mode~$d$ due to an outermost mode cast iff $s = \BFalse$
  \end{tabular}
  \begin{mathpar}
    \inferrule{x \in V}{V |-_{\Sigma, \Omega} \isTerm{x}^{\syn\,\BTrue\,\BTrue}}\,\SynRule{Var}
    \and
    \inferrule{\cdot |-_{\Sigma} A \\ V |-_{\Sigma, \Omega}\isTerm{t}^{\chk\,g\,s}}{V |-_{\Sigma, \Omega} (\isTerm{t \annotate A})^{\syn\,g\,\BTrue}}\,\SynRule{Anno}
    \and
    \inferrule{V |-_{\Sigma, \Omega}\isTerm{t}^{\chk\,g\,\BTrue}}{V |-_{\Sigma, \Omega} \isTerm{t}^{\syn\,\BFalse\,\BFalse}}\,\SynRule{Missing}
    \and
    \inferrule{V |-_{\Sigma, \Omega} \isTerm{t}^{\syn\,g\,\BTrue}}{V |-_{\Sigma, \Omega} \isTerm{t}^{\chk\,g\,\BFalse}}\,\ChkRule{Sub}
    \and
    \inferrule{V, \vec x_1 |-_{\Sigma, \Omega} \isTerm{t_1}^{\dir{d_1}\,g_1\,s_1} \\ \cdots \\ V, \vec x_n |-_{\Sigma, \Omega} \isTerm{t_n}^{\dir{d_n}\,g_n\,s_n}}
    {V |-_{\Sigma, \Omega} \tmOpts^{\dir{d}\,(\bigwedge_i g_i)\,\BTrue}}\,\Rule{Op}
  \end{mathpar}
  \caption{Generalised mode derivations}
  \label{fig:generalised-mode-derivations}
\end{figure}

@@

\begin{proof}
Induction on the given derivation.
The $\SynRule{Missing}$ rule cannot appear because $g = \BTrue$, and the other rules are mapped to their counterparts.
\end{proof}

@@

\begin{proof}[Proof of \cref{thm:Pre?-false}]
Induction on the generalised mode derivation, in each case analysing an arbitrary mode derivation and showing that it cannot exist.
The key case is $\SynRule{Missing}$, where we have a sub-derivation of $V |-_{\Sigma,\Omega} t^{\chk\,g\,\BTrue}$ for some boolean~$g$.
We do not have an induction hypothesis and seem to get stuck, because $g$~is not necessarily~$\BFalse$.
But what matters here is that $t$~is \emph{strictly} in checking mode: if we continue to analyse the sub-derivation, the outermost rule must be $\Rule{Op}$ with $\dir d = {\chk}$, implying that $t$~has to be an operation in checking mode.
Then a case analysis shows that it is impossible to have a (synthesising) mode derivation of $V |-_{\Sigma,\Omega} t^{\syn}$.
%\begin{itemize}
%\item $\SynRule{Anno}$:
%Any mode derivation must also use $\SynRule{Anno}$ as the outermost rule, and then the induction hypothesis suffices.
%\item $\ChkRule{Sub}$:
%We have a sub-derivation of $V |-_{\Sigma,\Omega} t^{\syn\,\BFalse\,\BTrue}$.
%Any mode derivation of $V |-_{\Sigma,\Omega} t^{\chk}$ must use $\ChkRule{Chk}$ as the outermost rule.
%(The $\Rule{Op}$ rule with $\dir d = {\chk}$ is impossible because the sub-derivation shows that $t$~is exactly a synthesising term.)
%We then have a sub-derivation of $V |-_{\Sigma,\Omega} t^{\syn}$, and can use the induction hypothesis.
%\end{itemize}
\end{proof}

@@

\begin{proof}
Case analysis on $\dir{d}$ and $\dir{d'}$, adding an outermost mode cast to change the given derivation to a different mode if $\dir{d} \neq \dir{d'}$.
Note that it is permissible to add an outermost mode cast because the antecedent~(\ref{eq:strict-derivation}) requires the given derivation to be strict.
\end{proof}

\begin{proof}[Proof of \cref{thm:generalised-mode-decoration}]
By \cref{thm:adjustment}, it suffices to prove statement~(\ref{eq:strict-derivation}) by induction on~$t$.

\begin{itemize}
\item $\Rule{Var}$ is mapped to $\SynRule{Var}$.
\item $\Rule{Anno}$:
Let $t = (t' \bbcolon A)$.
By the induction hypothesis and \cref{thm:adjustment}, there is a derivation of $V |-_{\Sigma,\Omega} {t'}^{\chk\,g\,s}$ for some booleans $g$~and~$s$, to which we apply $\SynRule{Anno}$.
\item $\Rule{Op}$:
Let $t = \tmOpts$.
By the induction hypotheses and \cref{thm:adjustment}, there is a derivation of $V, \vec x_i |-_{\Sigma,\Omega} {t_i}^{\dir{d_i}\,g_i\,s_i}$ for each sub-term~$t_i$; apply $\Rule{Op}$ to all the derivations.
\vspace{-\topsep-\baselineskip}
\end{itemize}
\end{proof}

@@

\begin{proof}
By \cref{thm:generalised-mode-decoration}, there is a derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,g\,s}$ for some booleans $g$~and~$s$, and then simply check~$g$ and apply either \cref{thm:Pre?-true} or \cref{thm:Pre?-false} to obtain $V |-_{\Sigma,\Omega} t^\dir{d}$ or $V |/-_{\Sigma,\Omega} t^\dir{d}$.
\end{proof}

@@

\begin{figure}
  \centering\small
  \judgbox{\isTerm{t} \sqsupseteq \isTerm{u}}{A raw term $t$ is more annotated than $u$ (for some bidirectional type system $(\Sigma, \Omega)$)}
  \begin{mathpar}
    \inferrule{\isTerm{t} \sqsupseteq \isTerm{u}}
              {(\isTerm{t} \annotate A) \sqsupseteq \isTerm{u}}\;\Rule{More}
    \and
    \inferrule{\vphantom{x : \Identifier}}
              {\isTerm{x} \sqsupseteq \isTerm{x}}
    \and
    \inferrule{\isTerm{t} \sqsupseteq \isTerm{u}}
              {(\isTerm{t} \annotate A) \sqsupseteq (\isTerm{u} \annotate A)}
    \and
    \inferrule{\isTerm{t_1} \sqsupseteq \isTerm{u_1} \quad \cdots \quad \isTerm{t_n} \sqsupseteq{u_n}}
              {\tmOpts \sqsupseteq \tmOpus}
  \end{mathpar}

@@

\begin{proof}[Proof of \cref{thm:annotatability}]
By \cref{thm:generalised-mode-decoration}, there is a generalised mode derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,g\,s}$ for some mode~$\dir{d}$ and booleans $g$~and~$s$.
Perform induction on this generalised mode derivation to construct a bidirectional typing derivation in the same mode (and also the term~$t'$ and a proof of $t' \sqsupseteq t$, which are determined by the derivations and omitted here).
\begin{itemize}
\item $\SynRule{Var}$, $\SynRule{Anno}$, and $\Rule{Op}$:
The outermost rule of the given typing derivation must be the corresponding one; apply the corresponding bidirectional typing rule to the induction hypotheses.
\item $\ChkRule{Sub}$:
Apply $\ChkRule{Sub}$ to the induction hypothesis.
\item $\SynRule{Missing}$:
This is the interesting case where we map a $\SynRule{Missing}$ rule to an $\SynRule{Anno}$ rule and add to the term a type annotation, which comes from the given typing derivation.
\vspace{-\topsep-\baselineskip}
\end{itemize}
\end{proof}

On the other hand, when using a bidirectional type synthesiser to implement type synthesis~(\cref{thm:implementation}), our completeness is much simpler to use than annotatability, which requires the bidirectional type synthesiser to produce more complex evidence when the synthesis fails.
Annotatability also does not help the user to deal with missing annotations like (generalised) mode decoration does: although annotatability seems capable of determining where annotations are missing and even filling them in correctly, its antecedent requires a typing derivation, which is what the user is trying to construct and does not have yet.
Therefore we believe that the notion of annotatability is too complex\LT{not clear what it means} for what it achieves, and our theory offers simpler and more useful alternatives.

>>>>

\section{Mode decoration and related properties}\label{sec:pre-synthesis}

@@

\subsection{Soundness and completeness}
\label{sec:soundness-and-completeness}

@@

%\begin{proof}
%From left to right:
%Induction on the bidirectional typing derivation, mapping every rule to its counterpart except when constructing a typing derivation from the $\ChkRule{Sub}$ rule, in which case the induction hypothesis $\Gamma |-_{\Sigma,\erase\Omega} t : B$ suffices due to the premise $B = A$.
%
%From right to left:
%Induction on the mode derivation.
%For \SynRule{Var}, \SynRule{Anno}, and \Rule{Op}, the outermost rule used in the typing derivation must be the corresponding typing rule, so by the induction hypotheses we have bidirectional typing derivations for all the sub-terms, to which we can then apply the corresponding bidirectional typing rule.
%The \ChkRule{Sub} case is similar but slightly simpler: the induction hypothesis directly gives us a derivation of $\Gamma |-_{\Sigma,\Omega} t \syn A$, to which we apply $\ChkRule{Sub}$.
%\end{proof}

\subsection{Generalised mode decoration}
\label{sec:mode-decoration}

@@

\begin{figure}
  \centering
  \small
  \begin{tabular}{ r r l }
    & & is in mode~$d$, \\
    $\smash{\boxed{V |-_{\Sigma, \Omega} \isTerm{t}^{\dir{d}\,g\,s}}}$
    & $\isTerm{t}$ \hspace{.1em}
    & misses some type annotation iff $g = \BFalse$, and \\
    & & is in mode~$d$ due to an outermost mode cast iff $s = \BFalse$
  \end{tabular}
  \begin{mathpar}
    \inferrule{x \in V}{V |-_{\Sigma, \Omega} \isTerm{x}^{\syn\,\BTrue\,\BTrue}}\,\SynRule{Var}
    \and
    \inferrule{\cdot |-_{\Sigma} A \\ V |-_{\Sigma, \Omega}\isTerm{t}^{\chk\,g\,s}}{V |-_{\Sigma, \Omega} (\isTerm{t \annotate A})^{\syn\,g\,\BTrue}}\,\SynRule{Anno}
    \and
    \inferrule{V |-_{\Sigma, \Omega}\isTerm{t}^{\chk\,g\,\BTrue}}{V |-_{\Sigma, \Omega} \isTerm{t}^{\syn\,\BFalse\,\BFalse}}\,\SynRule{Missing}
    \and
    \inferrule{V |-_{\Sigma, \Omega} \isTerm{t}^{\syn\,g\,\BTrue}}{V |-_{\Sigma, \Omega} \isTerm{t}^{\chk\,g\,\BFalse}}\,\ChkRule{Sub}
    \and
    \inferrule{V, \vec x_1 |-_{\Sigma, \Omega} \isTerm{t_1}^{\dir{d_1}\,g_1\,s_1} \\ \cdots \\ V, \vec x_n |-_{\Sigma, \Omega} \isTerm{t_n}^{\dir{d_n}\,g_n\,s_n}}
    {V |-_{\Sigma, \Omega} \tmOpts^{\dir{d}\,(\bigwedge_i g_i)\,\BTrue}}\,\Rule{Op}
  \end{mathpar}
  \caption{Generalised mode derivations}
  \label{fig:generalised-mode-derivations}
\end{figure}

@@

%\begin{proof}
%Induction on the given derivation.
%The $\SynRule{Missing}$ rule cannot appear because $g = \BTrue$, and the other rules are mapped to their counterparts.
%\end{proof}

@@

%\begin{proof}[Proof of \cref{thm:Pre?-false}]
%Induction on the generalised mode derivation, in each case analysing an arbitrary mode derivation and showing that it cannot exist.
%The key case is $\SynRule{Missing}$, where we have a sub-derivation of $V |-_{\Sigma,\Omega} t^{\chk\,g\,\BTrue}$ for some boolean~$g$.
%We do not have an induction hypothesis and seem to get stuck, because $g$~is not necessarily~$\BFalse$.
%But what matters here is that $t$~is \emph{strictly} in checking mode: if we continue to analyse the sub-derivation, the outermost rule must be $\Rule{Op}$ with $\dir d = {\chk}$, implying that $t$~has to be an operation in checking mode.
%Then a case analysis shows that it is impossible to have a (synthesising) mode derivation of $V |-_{\Sigma,\Omega} t^{\syn}$.
%%\begin{itemize}
%%\item $\SynRule{Anno}$:
%%Any mode derivation must also use $\SynRule{Anno}$ as the outermost rule, and then the induction hypothesis suffices.
%%\item $\ChkRule{Sub}$:
%%We have a sub-derivation of $V |-_{\Sigma,\Omega} t^{\syn\,\BFalse\,\BTrue}$.
%%Any mode derivation of $V |-_{\Sigma,\Omega} t^{\chk}$ must use $\ChkRule{Chk}$ as the outermost rule.
%%(The $\Rule{Op}$ rule with $\dir d = {\chk}$ is impossible because the sub-derivation shows that $t$~is exactly a synthesising term.)
%%We then have a sub-derivation of $V |-_{\Sigma,\Omega} t^{\syn}$, and can use the induction hypothesis.
%%\end{itemize}
%\end{proof}

@@

%\begin{proof}
%Case analysis on $\dir{d}$ and $\dir{d'}$, adding an outermost mode cast to change the given derivation to a different mode if $\dir{d} \neq \dir{d'}$.
%Note that it is permissible to add an outermost mode cast because the antecedent~(\ref{eq:strict-derivation}) requires the given derivation to be strict.
%\end{proof}

\begin{proof}[\cref{thm:generalised-mode-decoration}]
By \cref{thm:adjustment}, it suffices to prove~\eqref{eq:strict-derivation} by induction on~$t$.

\begin{itemize}
\item $\Rule{Var}$ is mapped to $\SynRule{Var}$.
\item $\Rule{Anno}$:
Let $t = (t' \bbcolon A)$.
By the induction hypothesis and \cref{thm:adjustment}, there is a derivation of $V |-_{\Sigma,\Omega} {t'}^{\chk\,g\,s}$ for some booleans $g$~and~$s$, to which we apply $\SynRule{Anno}$.
\item $\Rule{Op}$:
Let $t = \tmOpts$.
By the induction hypotheses and \cref{thm:adjustment}, there is a derivation of $V, \vec x_i |-_{\Sigma,\Omega} {t_i}^{\dir{d_i}\,g_i\,s_i}$ for each sub-term~$t_i$; apply $\Rule{Op}$ to all the derivations.
\end{itemize}
\end{proof}

@@

%\begin{proof}
%By \cref{thm:generalised-mode-decoration}, there is a derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,g\,s}$ for some booleans $g$~and~$s$, and then simply check~$g$ and apply either \cref{thm:Pre?-true} or \cref{thm:Pre?-false} to obtain $V |-_{\Sigma,\Omega} t^\dir{d}$ or $V |/-_{\Sigma,\Omega} t^\dir{d}$.
%\end{proof}

@@

\begin{figure}
  \centering\small
  \judgbox{\isTerm{t} \sqsupseteq \isTerm{u}}{A raw term $t$ is more annotated than $u$}
  \begin{mathpar}
    \inferrule{\isTerm{t} \sqsupseteq \isTerm{u}}
              {(\isTerm{t} \annotate A) \sqsupseteq \isTerm{u}}\;\Rule{More}
    \and
    \inferrule{\vphantom{x : \Identifier}}
              {\isTerm{x} \sqsupseteq \isTerm{x}}
    \and
    \inferrule{\isTerm{t} \sqsupseteq \isTerm{u}}
              {(\isTerm{t} \annotate A) \sqsupseteq (\isTerm{u} \annotate A)}
    \and
    \inferrule{\isTerm{t_1} \sqsupseteq \isTerm{u_1} \quad \cdots \quad \isTerm{t_n} \sqsupseteq{u_n}}
              {\tmOpts \sqsupseteq \tmOpus}
  \end{mathpar}

@@

\begin{proof}[Proof of \cref{thm:annotatability}]
By \cref{thm:generalised-mode-decoration}, there is a generalised mode derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,g\,s}$ for some mode~$\dir{d}$ and booleans $g$~and~$s$.
Perform induction on this generalised mode derivation to construct a bidirectional typing derivation in the same mode (and also the term~$t'$ and a proof of $t' \sqsupseteq t$, which are determined by the derivations and omitted here).
\begin{itemize}
\item $\SynRule{Var}$, $\SynRule{Anno}$, and $\Rule{Op}$:
The outermost rule of the given typing derivation must be the corresponding one; apply the corresponding bidirectional typing rule to the induction hypotheses.
\item $\ChkRule{Sub}$:
Apply $\ChkRule{Sub}$ to the induction hypothesis.
\item $\SynRule{Missing}$:
This is the interesting case where we map a $\SynRule{Missing}$ rule to an $\SynRule{Anno}$ rule and add to the term a type annotation, which comes from the given typing derivation.
\end{itemize}
\end{proof}

On the other hand, when using a bidirectional type synthesiser to implement type synthesis~(\cref{thm:implementation}), our completeness is much simpler to use than annotatability, which requires the bidirectional type synthesiser to produce more complex evidence when the synthesis fails.
Annotatability also does not help the user to deal with missing annotations like (generalised) mode decoration does: although annotatability seems capable of determining where annotations are missing and even filling them in correctly, its antecedent requires a typing derivation, which is what the user is trying to construct and does not have yet.
%Therefore we believe that the notion of annotatability is too complex\LT{not clear what it means} for what it achieves, and our theory offers simpler and more useful alternatives.
